Adaptive evolution with aneuploidy and mutation

### 2021-12-26 ms/ms.tex
following phone call between YR and IK

\pagebreak
% Models and Methods
\section*{Models and Methods}

%%%%%%%% Fig illustation
\begin{figure}[b!]
  %\centering
  \begin{subfigure}[t]{0.4\textwidth}
      \caption{
        \textbf{A. Single-locus model}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}      
      \label{fig:model1}
  \end{subfigure}
  %\\
  \begin{subfigure}[t]{0.4\textwidth}
  	  \caption{
        \textbf{B. Multi-locus model}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-B.pdf}      
      \label{fig:model2}
  \end{subfigure}
  \caption{
    \textbf{Model illustrations.}
    \textbf{(A)} In the single-locus model, the four genotypes are: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
    Overall there are two possible trajectories from \euwt\ to \eumt.
    \textbf{(B)} In the multi-locus model, each genotype is characterized by its karyotype, \euwt\ or \anwt, and the number of accumulated beneficial mutations, denoted by stars. 
    In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$ and $\tau \mu$, beneficial mutation rates in euploid and aneuploid cells; $\delta$, aneuploidy gain rate; $\delta_L$, aneuploidy loss rate.
    % TODO IK: change k mu to tau mu in figure; also k mu is too close to the arrow right now; change alpha to delta in the figure 
  }
  \label{fig:models}
\end{figure}

\paragraph*{Evolutionary Models.}
We model the evolution of a population of cells using two models: a single-locus model and a multi-locus model.
Both models are based on the Wright-Fisher model~\citep{Otto2007}, assuming non-overlapping generations and including the effects of natural selection, genetic drift, aneuploidy, and mutation. 
We focus on beneficial mutations, neglecting the effects of deleterious and neutral mutations. Both models allow for a single aneuploid karyotype (e.g., chromosome III duplication).
While the single-locus model allows for only a single mutation to accumulate in the genotype, the multi-locus model allows for multiple mutations to accumulate~(\autoref{fig:models}), as well as for a fluctuating population size.

\paragraph*{Single-locus model.}
This model assumes a constant effective population size $N$ and follows four genotypes~(\autoref{fig:models}A): euploid wild-type, \euwt, the initial genotype; 
euploid mutant, \eumt, with the standard karyotype and a single beneficial mutation; 
aneuploid wild-type, \anwt, with an extra chromosome, i.e., following chromosome duplication; and
aneuploid mutant, \anmt, with and extra chromosome and a beneficial mutation.

Transitions between the genotypes occur as follows~(\autoref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; occur with probability $\mu$, the mutation rate, and from \anwt\; to \anmt\; with probability $\tau \mu$, where $\tau$ is the fold-change in the rate of beneficial mutations in aneuploid cells. By default, we assume $\tau=33/32$, as the wild-type \yeast strains in the experiments by \citep{Yona2012} are diploid with 32 chromosomes, and the aneuploid strains are trisomic, with 33 chromosomes.
Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\ with probability $\delta$, the aneuploidy gain rate.
Aneuploidy is lost, transitioning cells from \anmt\; to \eumt\; with probability $\delta_L$, the aneuploidy loss rate.
The fitness values of the four genotypes are given by \autoref{table:single-locus}.

%%% Table: single-locus model
\begin{table}[h]
\centering
\caption{\textbf{Single-locus model fitness values.}}
\begin{tabular}{lllll}
\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^*$ & $2n^*$ \\
\hline
\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)(1+s)+b$ & $1+s$               
\end{tabular}
\label{table:single-locus}
\caption*{
$s \ge 0$ is the selection coefficient of a beneficial mutation;
$b \ge 0$ is the selection coefficient of aneuploidy; 
and $0 \le c \le 1$ is the fitness cost of aneuploidy.
% TODO IK do we assume b>c?
% IK ANS we assume that 2n+1>2n, see below, so yes.
}
\end{table}
%%%

The initial population has $N$ cells with genotype \euwt.
The effect of natural selection on the frequency $f_i$ of genotype $i = 2n, 2n+1, 2n+1^*, \text{or } 2n^*$ is given by
    \begin{equation} \label{eq:selection-single} 
      f^s_i = \frac{f_i w_i}{\bar{w}} \;,
    \end{equation}
where the fitness values $w_i$ are given in \autoref{table:single-locus} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
The effect of mutation and aneuploidy on genotype frequencies is given by
% TODO IK we are missing 2n+1->2n, 2n*->2n+1*, why? to neglect them we need to assume w(2n+1)>w(2n) and w(2n*)>w(2n+1*), but I thought we don't assume that but rather we infer that. 
% IK ANS We know that w(2n+1)>w(2n) because trisomy is fixated in the experiment. We can also assume that w(2n*)>w(2n+1) by Fig4A of Yona paper (refined vs trisomic growth) and because it is fixated. 
% Actually I assume in code that  w(2n*)>w(2n+1*)>w(2n+1)>w(2n) just for speeding up the simulations (I don't check some parameters combinations in this way). But I checked also that when I don't adding this constrain the result was the same. However I dind't add reverse direction in this case.
    \begin{equation} \label{eq:mutation-aneuploidy-single}
    \begin{aligned}
      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n}  \;,\\
      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \tau \mu) f^s_{2n+1}  \;,\\
      &f^m_{2n+1^*} &=&\; \tau \mu f^s_{2n+1} + (1-\delta_L) f^s_{2n+1^*}  \;,\\
      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta_L f^s_{2n+1} + f^s_{2n^*}  \;.
    \end{aligned}
    \end{equation}
Finally, random genetic drift is modeled using a multinomial distribution~\citep{Otto2007},
    \begin{equation} \label{eq:drift-single}
      \vec{f'} \sim \frac{1}{N} \cdot \mathit{Mult}(N,\ \vec{f^m}) \;,
    \end{equation}
where $\vec{f^m}=(f^m_{2n}\ ,\ f^m_{2n+1}\ ,\ f^m_{2n+1^*}\ ,\ f^m_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N,\ \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.

\paragraph*{Multi-locus model.}
This model expands the single-locus model by allowing for (i) the accumulation of beneficial mutations, and (ii) a fluctuating population size.

A genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, which can be zero or more.
The selection coefficient of the $i$-th accumulated mutation in each individual, $s_i$, is drawn from an exponential distribution with expected value $s$: $s_i \sim \mathit{Exp{(s)}}$.
The rest of the parameters ($N$, $\mu$, $\tau$, $\delta$, $\delta_L$, $b$, $c$) are the same as in the single-locus model.
% ILIA: I don’t know if it should be mentioned here, but the meaning of the mutation rate is a bit differ for multi-locus model. For the single-locus mutation rate value is the probability to get specific mutation (maybe one of a few). For the multi-locus it is the probability to get some beneficial mutation (one of a many). So, we don’t expect to receive same values for mutation rate for the models.
% YOAV: let's talk about it, I don't necessarily agree
% ILIA: I don’t have in the code 𝜏 parameter for multi-locus model (it’s like 1 there). Do you think it worth adding it for consistency?
% YOAV: if our inference tells us that tau=1 is as good as tau>1, then maybe we can remove this parameter from the model presentation altogether, and simply introduce it when we discuss this result as part of the sensitivity analysis section
The fitness of the different genotypes is the same as in the single-locus model (\autoref{table:single-locus}), except that the fitness contribution of $k$ beneficial mutations is the product of their independent effects, $\prod_{i=1}^k(1+s_i)$, instead of the contribution of the single mutation allowed in the single-locus model, $(1+s)$, see \autoref{table:multi-locus}.
Therefore, aneuploidy loss would be favored by selection only if there are enough beneficial mutations and/or the selection coefficients $s_i$ are large enough. The intuition is that when the benefit of the accumulated beneficial mutations is small, ten the benefit of aneuploidy has a large effect; when the benefit of the accumulated beneficial mutations is large, then aneuploidy is no longer beneficial due of its fitness cost.

@@

The effects of mutation and aneuploidy on genotype frequencies is more elaborate than in the single-locus model.
Genotype $i$ is classified according to its karyotype (\euwt\ or \anwt), the number of accumulated beneficial mutations ($k \ge 0$), and their fitness ($w_i$).
Each offspring cell inherits these properties from its mother cell.
Then, with probability $\mu$ or $\tau \mu$ for euploid and aneuploid cells, % TODO no tau in multi-locus model?
respectively, a new beneficial mutation is accumulated, such that the number of mutations is $k+1$, and its effect $s_{k+1}$ is drawn from an exponential distribution with expected value $s$, such that the contribution of the mutations to the fitness is $\prod_{j=0}^{k+1}{(1+s_j)}$.
Next, euploid offspring become aneuploid with probability $\delta$, and aneuploid offspring become euploid with probability $\delta_L$.

%%% Table: multi-locus model
\begin{table}[h]
\centering
\caption{\textbf{Multi-locus model fitness values.}}
\begin{tabular}{lllll}
\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^{*k}$ & $2n^{*k}$ \\
\hline
\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)\prod_{j=1}^k(1+s_i)+b$ & $\prod_{j=1}^k(1+s_i)$               
\end{tabular}
\label{table:multi-locus}
\caption*{
$k$ is the number of accumulated beneficial mutations;
$s \ge 0$ is the selection coefficient of a beneficial mutation;
$b \ge 0$ is the selection coefficient of aneuploidy; 
and $0 \le c \le 1$ is the fitness cost of aneuploidy.
%TODO do we assume b>c ?
}
\end{table}
%%%

\paragraph{Empirical evidence.}
We use empirical results of evolutionary experiments reported by \citet{Yona2012}.
In their heat-stress experiment, four populations of \yeast evolved under \SI{39}{\celsius}. Aneuploidy fixed in all four population in the first 450 generations (hereafter, fixation or elimination of a genotype \emph{by generation $t$} means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier). From unpublished results, aneuploidy did not fix before at least 200 generations elapsed.
The experiment continued with two populations, in which aneuploidy was eliminated by generation 1700 and 2350.
%In the pH-stress experiment, four populations of \yeast evolved under high-ph stress (8.6). Aneuploidy fixed during the first 150 generations. It was fully eliminated in two populations and partly eliminated in two populations by generation 750.
%TODO do we use pH?

\paragraph{Likelihood function.}
Denote by $A_{t,T}$ the fixation of aneuploidy by generation $t$ conditioned on no fixation up to generation $T$; by $B_{t}$ the extinction of aneuploidy by generation $t$; and by $C_{t,T}$ the extinction of aneuploidy by generation $t$ conditioned on no extinction by generation $T$.
The model likelihood function for parameter vector $\theta$ for the heat-stress experiment is
\begin{multline} \label{eq:heatstress-likelihood}
\likelihood(\theta\ |\ A_{450,200}\ ,\ B_{1700}\ ,\ C_{2350, 1700} ) = \\
P^4(A_{450,200}) \cdot \Big(1 - P^4(\lnot{B_{1700}} | A_{450,200} ) - P^4(\lnot{C_{2350, 1700}}| A_{450,200}) + P^4(\lnot{B_{1700}} \land \lnot{L^*_{2350, 1700}}| A_{450,200})\Big)	\;,	
\end{multline}
where $P(X)$ is the probability of event $X$ and $\lnot$ means \emph{not}.

%The likelihood function of the model with parameter set $\theta$ for the pH-stress experiment is
%\begin{equation} \label{eq:phstress-likelihood} % TODO: add ph-Stress model to results or remove it from here (if remained add A_t to notations, it is At,T now only)
%\likelihood(\theta | A_{150}, L_{750} ) = \\
% P^4(B_{150}) \cdot 6 \cdot	P^2(B_{600} | A_{150}) \cdot P^2(\lnot{B_{600}} | A_{150}) \;.  
% \end{equation}

For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), denote by $M_{t}$ the fixation of \eumt\ by generation $t$, and by $Q_{t,T}$ the fixation of \eumt\ by generation $t$ conditioned on no fixation at generation $T$.
The likelihood of a model without aneuploidy with parameter vector $\theta$ for the heat-stress experiment is
\begin{multline} \label{eq:heatstress-noaneuploidy-likelihood}
\likelihood(\theta |  M_{1700}, M^*_{2350, 1700} ) = 
\Big(1 - P^4(\lnot{M_{1700}}) - P^4(\lnot{Q_{2350, 1700}}) + P^4(\lnot{M_{1700}} \land \lnot{Q_{2350, 1700}})\Big)	\;,	
\end{multline}

Because our model, just like the Wright-Fisher model, is non-linear and stochastic, these likelihood functions are intractable. We overcome this problem by approximating the likelihood using multiple simulations (1,000 simulations per parameter vector for the single-locus model). Denote by $\hat{\likelihood}(X | C)$ the approximate likelihood based on data of simulations $X$ and the expected conditions $C$.
%TODO if we have more models, write how much replicates were used.
% TODO ref to place where we show 1,000 is enough

\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation~\citep{Sunnaker2013} with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyABC} Python 
package~\citep[\href{https://pyabc.readthedocs.io}{pyabc.readthedocs.io}]{Klinger2018}. This approach uses numerical stochastic simulations of the model to infer a posterior distribution over the model parameters. It is a method of likelihood-free, simulation-based inference~\citep{Cranmer2020}, that is, for estimating a posterior distribution when a likelihood function cannot be directly computed. It is therefore suitable to our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed. 
%a bit adapted/expanded description from Syga et al. 2021: % TODO if you adapt from it, then we should cite it.

Briefly, the ABS-SMC algorithm employs sequential importance sampling over multiple iterations. 
In iteration $t$, the algorithm draws parameter vectors $\theta_i$, also called \emph{particles}, from a given proposal distribution and simulates data $X_i$ from the model until $n_t$ particles are accepted. Acceptance is based on a comparison to observed data $X_{obs}$ using a distance function $\Delta(X_i, X_{obs})$ and acceptance threshold $\epsilon_t$, such that $\Delta(X_i, X_{obs}) \leq \epsilon_t$. %TODO cite Syga here? 
In our case $X_i$ is 1000 %TODO check if this is the number for all models
simulations of the model, and the distance function is $1-\hat{\likelihood}(X_i | X_{obs})$
% TODO something about how the weight is computed and used?
The acceptance threshold $\epsilon_t$ is chosen as the median of the distances of the accepted particles in the previous iteration and $\epsilon_0=0.01$. 
When $n_t$ particles are accepted, a new proposal distribution is constructed using a multivariate normal kernel with adaptive covariance matrix. % TODO ref
In the first iteration the prior (see below) is used as a proposal distribution. 
$n_t$ is adapted at every iteration $t$ using the \emph{adaptive population strategy}~\citep{Klinger2017}.
The algorithm terminates when no more sufficient change in acceptance threshold occur, and no change in posterior. % TODO unclear
Inference with the single-locus model terminated after four iterations with $\epsilon=0.13$, 1074 accepted parameter vectors and effective sample size (ESS) 700 (\autoref{fig:ess-single}). Running the inference algorithm with different initialization seeds and different number of simulations with produced similar outcomes (\autoref{fig:seeds}). %TODO if we have more models, write about them here also.

After producing a set of weighted particles from the the posterior distribution using the above ABC-SMC algorithm, we approximate the posterior using kernel density estimation (KDE) with Gaussian kernels. We then sample $50,000$ samples from the posterior KDE to compute the posterior mode, highest density interval (HDI), plotting the posterior, etc.

\paragraph{Model comparison.} We perform model selection using WAIC, the widely applicable information criterion ~\citep{gelman2013bayesian},
\begin{equation} \label{eq:WAIC}
\begin{aligned}
\mathit{WAIC}(\theta, \vec{X}) &\ =\ 
-2\log\mathbb{E}[\mathcal{L}(\theta \mid \vec{X})]\ +\ 2\mathbb{V}[\log\mathcal{L}(\theta \mid \vec{X})]
\end{aligned}
\end{equation}
where $\theta$ is a parameter vector, $X$ are the model or experiment results (i.e., fixation of aneuploidy by generation $t$ etc.), and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution $P(\theta \mid \vec{X})$.
WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy and a difference of 2 is a popular threshold for model comparison~\citep{Kass1995}.
% TODO do we need to say "and a difference of 2 is a popular threshold for model comparison."? Meanwile WAICs are so small that there's no such difference
% YOAV: let's discuss
 % TODO do we use more methods? Describe probably posterior plots comparison method, and Bayes factor if it will used.

>>>>

\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%
% Models and Methods
\section*{Models and Methods}

%%%%%%%% Fig illustation
\begin{figure}[b!]
  %\centering
  \begin{subfigure}[t]{0.4\textwidth}
      \caption{
        \textbf{A. Single-locus model}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}      
      \label{fig:model1}
  \end{subfigure}
  %\\
  \begin{subfigure}[t]{0.4\textwidth}
  	  \caption{
        \textbf{B. Multi-locus model}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-B.pdf}      
      \label{fig:model2}
  \end{subfigure}
  \caption{
    \textbf{Model illustrations.}
    \textbf{(A)} In the single-locus model, the four genotypes are: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
    Overall there are two possible trajectories from \euwt\; to \eumt.
    \textbf{(B)} In the multi-locus model, each genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, denoted by stars. 
    In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$ and $\tau \mu$, beneficial mutation rates in euploid and aneuploid cells; $\delta$, aneuploidy rate.
    % TODO IK: change k mu to tau mu in figure; also k mu is too close to the arrow right now; change alpha to delta in the figure 
  }
  \label{fig:models}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Evolutionary Models.}
We model the evolution of a population of cells using two models: a single-locus model and a multi-locus model.
Both models are based on the Wright-Fisher model~\citep{Otto2007}, assuming non-overlapping generations and including the effects of natural selection, genetic drift, aneuploidy, and mutation. 
We focus on beneficial mutations, neglecting the effects of deleterious and neutral mutations. Both models allow for a single aneuploid karyotype (e.g., chromosome III duplication).
While the single-locus model allows for only a single mutation to accumulate in the genotype, the multi-locus model allows for multiple mutations to accumulate~(\autoref{fig:models}), as well as for a fluctuating population size.

%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Single-locus model.}
This model assumes a constant effective population size $N$ and follows four genotypes~(\autoref{fig:models}A): euploid wild-type, \euwt, the initial genotype; 
euploid mutant, \eumt, with the standard karyotype and a single beneficial mutation; 
aneuploid wild-type, \anwt, with an extra chromosome, i.e., following chromosome duplication; and
aneuploid mutant, \anmt, with and extra chromosome and a beneficial mutation.

Transitions between the genotypes occur as follows~(\autoref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; occur with probability $\mu$, the mutation rate, and from \anwt\; to \anmt\; with probability $\tau \mu$, where $\tau$ is the fold-change in the rate of beneficial mutations in aneuploid cells. By default, we assume $\tau=33/32$, as the wild-type \yeast strains in the experiments by \citep{Yona2012} are diploid with 32 chromosomes, and the aneuploid strains are trisomic, with 33 chromosomes. We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\; and from \anmt\ to \eumt\ with probability $\delta$, the aneuploidy rate. That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
The fitness values of the four genotypes are given by \autoref{table:single-locus}.

%%% Table: single-locus model
\begin{table}[h]
\centering
\caption{\textbf{Single-locus model fitness values.}}
\begin{tabular}{lllll}
\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^*$ & $2n^*$ \\
\hline
\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)(1+s)+b$ & $1+s$               
\end{tabular}
\label{table:single-locus}
\caption*{
$s \ge 0$ is the selection coefficient of a beneficial mutation;
$0 \le c \le 1$ is the fitness cost of aneuploidy;
and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%

The initial population has $N$ cells with genotype \euwt.
The effect of natural selection on the frequency $f_i$ of genotype $i = 2n, 2n+1, 2n+1^*, \text{or } 2n^*$ is given by
    \begin{equation} \label{eq:selection-single} 
      f^s_i = \frac{f_i w_i}{\bar{w}} \;,
    \end{equation}
where the fitness values $w_i$ are given in \autoref{table:single-locus} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
The effect of mutation and aneuploidy on genotype frequencies is given by
    \begin{equation} \label{eq:mutation-aneuploidy-single}
    \begin{aligned}
      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n}  \;,\\
      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \tau \mu) f^s_{2n+1}  \;,\\
      &f^m_{2n+1^*} &=&\; \tau \mu f^s_{2n+1} + (1-\delta) f^s_{2n+1^*}  \;,\\
      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1} + f^s_{2n^*}  \;.
    \end{aligned}
    \end{equation}
Finally, random genetic drift is modeled using a multinomial distribution~\citep{Otto2007},
    \begin{equation} \label{eq:drift-single}
      \vec{f'} \sim \frac{1}{N} \cdot \mathit{Mult}(N,\ \vec{f^m}) \;,
    \end{equation}
where $\vec{f^m}=(f^m_{2n}\ ,\ f^m_{2n+1}\ ,\ f^m_{2n+1^*}\ ,\ f^m_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N,\ \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.

%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Multi-locus model.}
This model expands the single-locus model by allowing for (i) the accumulation of beneficial mutations, and (ii) a fluctuating population size.

A genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, which can be zero or more.
The selection coefficient of the $i$-th accumulated mutation in each individual, $s_i$, is drawn from an exponential distribution with expected value $s$: $s_i \sim \mathit{Exp{(s)}}$.
The rest of the parameters ($N$, $\mu$, $\tau$, $\delta$, $\delta_L$, $b$, $c$) are the same as in the single-locus model. However, since the multi-locus model allows several mutations with smaller fitness effects to accumulate, we expect the mutation rate to be higher compared to the single-locus model, which focuses on a single, large-effect mutation.
% TODO : add tau to model fixed at 33/32.

The fitness of the different genotypes is the same as in the single-locus model (\autoref{table:single-locus}), except that the fitness contribution of $k$ beneficial mutations is the product of their independent effects, $\prod_{i=1}^k(1+s_i)$, instead of the contribution of the single mutation allowed in the single-locus model, $(1+s)$, see \autoref{table:multi-locus}.
Therefore, aneuploidy loss would be favored by selection only if there are enough beneficial mutations and/or the selection coefficients $s_i$ are large enough. The intuition is that when the benefit of the accumulated beneficial mutations is small, ten the benefit of aneuploidy has a large effect; when the benefit of the accumulated beneficial mutations is large, then aneuploidy is no longer beneficial due of its fitness cost.

@@

The effects of mutation and aneuploidy on genotype frequencies is more elaborate than in the single-locus model.
Genotype $i$ is classified according to its karyotype (\euwt\ or \anwt), the number of accumulated beneficial mutations ($k \ge 0$), and their fitness ($w_i$).
Each offspring cell inherits these properties from its mother cell.
Then, with probability $\mu$ or $\tau \mu$ for euploid and aneuploid cells,
respectively, a new beneficial mutation is accumulated, such that the number of mutations is $k+1$, and its effect $s_{k+1}$ is drawn from an exponential distribution with expected value $s$, such that the contribution of the mutations to the fitness is $\prod_{j=0}^{k+1}{(1+s_j)}$.
Next, euploid offspring become aneuploid with probability $\delta$, and aneuploid offspring become euploid with probability $\delta_L$.

%%% Table: multi-locus model
\begin{table}[h]
\centering
\caption{\textbf{Multi-locus model fitness values.}}
\begin{tabular}{lllll}
\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^{*k}$ & $2n^{*k}$ \\
\hline
\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)\prod_{j=1}^k(1+s_i)+b$ & $\prod_{j=1}^k(1+s_i)$               
\end{tabular}
\label{table:multi-locus}
\caption*{
$k$ is the number of accumulated beneficial mutations;
$s \ge 0$ is the selection coefficient of a beneficial mutation;
$0 \le c \le 1$ is the fitness cost of aneuploidy;
and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
% TODO check if b > c in code
}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

\paragraph{Empirical evidence.}
We use empirical results of evolutionary experiments reported by \citet{Yona2012}.
In their heat-stress experiment, four populations of \yeast evolved under \SI{39}{\celsius}. Aneuploidy fixed in all four population in the first 450 generations (hereafter, fixation or elimination of a genotype \emph{by generation $t$} means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier). From unpublished results, aneuploidy did not fix before at least 200 generations elapsed.
The experiment continued with two populations, in which aneuploidy was eliminated by generation 1,700 and 2,350.

\paragraph{Likelihood function.} % TODO fix this .... (YR)
Denote by $A_{t,T}$ the fixation of aneuploidy by generation $t$ conditioned on no fixation up to generation $T$; by $B_{t}$ the extinction of aneuploidy by generation $t$; and by $C_{t,T}$ the extinction of aneuploidy by generation $t$ conditioned on no extinction by generation $T$.
The model likelihood function for parameter vector $\theta$ for the heat-stress experiment is
\begin{multline} \label{eq:heatstress-likelihood}
\likelihood(\theta) = \\
P^4(A_{450,200}) \cdot \Big(1 - P^4(\lnot{B_{1700}} | A_{450,200} ) - P^4(\lnot{C_{2350, 1700}}| A_{450,200}) + P^4(\lnot{B_{1700}} \land \lnot{L^*_{2350, 1700}}| A_{450,200})\Big)	\;,	 % TODO fix L* (YR)
\end{multline}
where $P(X)$ is the probability of event $X$ and $\lnot$ means \emph{not}.

For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), denote by $M_{t}$ the fixation of \eumt\ by generation $t$, and by $Q_{t,T}$ the fixation of \eumt\ by generation $t$ conditioned on no fixation at generation $T$.
The likelihood of a model without aneuploidy with parameter vector $\theta$ for the heat-stress experiment is
\begin{multline} \label{eq:heatstress-noaneuploidy-likelihood}
\likelihood_{!}(\theta) = 
\Big(1 - P^4(\lnot{M_{1700}}) - P^4(\lnot{Q_{2350, 1700}}) + P^4(\lnot{M_{1700}} \land \lnot{Q_{2350, 1700}})\Big)	\;,	
\end{multline}

Because our model, just like the Wright-Fisher model, is non-linear and stochastic, these likelihood functions are intractable. We overcome this problem by approximating the likelihood using simulations: we simulate 1,000 experiments per parameter vector. This approximate likelihood is denoted by $\hat{\likelihood}$.
% TODO if we show results of multi-locus model, write how much replicates were used.
% TODO ref to place where we show 1,000 is enough

\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation~\citep{Sunnaker2013} with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyABC} Python 
package~\citep[\href{https://pyabc.readthedocs.io}{pyabc.readthedocs.io}]{Klinger2018}. This approach uses numerical stochastic simulations of the model to infer a posterior distribution over the model parameters. It is a method of likelihood-free, simulation-based inference~\citep{Cranmer2020}, that is, for estimating a posterior distribution when a likelihood function cannot be directly computed. It is therefore suitable to our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed.

Briefly, the ABS-SMC algorithm employs sequential importance sampling over multiple iterations~\citep{Syga2021}. 
In iteration $t$, the algorithm draws parameter vectors $\theta_i$, also called \emph{particles}, from a given proposal distribution and simulates data $X_i$ from the model until $n_t$ particles are accepted. 
We accept a proposal $\theta_i$ if its approximate likelihood $\hat{\likelihood}(\theta_i)$
Acceptance is based on a comparison to observed data $X_{obs}$ using a distance function $\Delta(X_i, X_{obs})$ and acceptance threshold $\epsilon_t$, such that $\Delta(X_i, X_{obs}) \leq \epsilon_t$.  % TODO replace Delta with approx likelihood (YR)
In our case $X_i$ is 1,000 simulations of the model, and the distance function is $1-\hat{\likelihood}(X_i | X_{obs})$
% TODO something about how the weight is computed and used?
The acceptance threshold $\epsilon_t$ is chosen as the median of the distances of the accepted particles in the previous iteration and $\epsilon_0=0.01$. 
When $n_t$ particles are accepted, a new proposal distribution is constructed using a multivariate normal kernel with adaptive covariance matrix. % TODO ref - Klinger? Syga? Sisson?
In the first iteration the prior (see below) is used as a proposal distribution. 
$n_t$ is adapted at every iteration $t$ using the \emph{adaptive population strategy}~\citep{Klinger2017}.
Acceptance is determined according to the approximate likelihood, which has a maximum value of 0.875. Thus, we terminated the inference when $\epsilon \le 0.13$ after four iterations, with 1,074 accepted parameter vectors and effective sample size (ESS) 700 (\autoref{fig:ess-single}). % TODO for multi-locus model we need to say something else
Running the inference algorithm with different initialization seeds and different number of simulations with produced similar outcomes (\autoref{fig:seeds}). %TODO for multi-locus model, write about them here also.

After producing a set of weighted particles from the the posterior distribution using the above ABC-SMC algorithm, we approximate the posterior using kernel density estimation (KDE) with Gaussian kernels, from which we find the MAP (maximum a posteriori) estimate as the maximum of the KDE function. We then sample $50,000$ samples from the posterior KDE to compute the HDI (highest density interval) and visualize the posterior distribution with histograms.

\paragraph{Model comparison.} We perform model selection using WAIC, the widely applicable information criterion ~\citep{gelman2013bayesian},
\begin{equation} \label{eq:WAIC}
\begin{aligned}
\mathit{WAIC}(\theta, \vec{X}) &\ =\ 
-2\log\mathbb{E}[\mathcal{L}(\theta \mid \vec{X})]\ +\ 2\mathbb{V}[\log\mathcal{L}(\theta \mid \vec{X})]
\end{aligned}
\end{equation}
where $\theta$ is a parameter vector, $X$ are the model or experiment results (i.e., fixation of aneuploidy by generation $t$ etc.), and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution $P(\theta \mid \vec{X})$.
WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy and a difference of 2 is a popular threshold for model comparison~\citep{Kass1995}.
% TODO Describe posterior plots comparison method - Fig 4 & S3
